Storm surges in the Strait of Georgia simulated with a regional model

### 2015-05-14 AO_guides/manuscriptAO.tex
Checked consistency up until Nov 2006

\begin{abstract}
The Strait of Georgia is a large, semi-enclosed body of water between Vancouver Island and the mainland of British Columbia connected to the Pacific Ocean via the Strait of Juan de Fuca at the south and Johnstone Strait at the north. During the winter months, coastal communities in the Strait of Georgia are at risk to flooding caused by storm surges, a natural hazard that can occur when a strong storm coincides with high tide. This investigation produces storm surge hindcasts using a three-dimensional numerical ocean model for the Strait of Georgia and the surrounding bodies of water (Strait of Juan de Fuca, Puget Sound, and Johnstone Strait) collectively known as the Salish Sea. The numerical model employs the Nucleus for European Modelling of the Ocean (NEMO) architecture in a regional configuration. The model is evaluated through {\color{red} the} comparisons of tidal elevation harmonics and storm surge with observations. Important forcing factors contributing to storm surges have been assessed. It is shown that the surge entering the domain from the Pacific Ocean contributes most significantly to the {\color{red} surge amplitude within the Strait of Georgia. Comparisons between simulations with high resolution and low resolution atmospheric forcing further emphasize tthat remote forcing is the dominant factor in surge amplitudes of this region.} In addition, local wind patterns cause a slight increase in surge amplitude on the mainland side of the Strait of Georgia when compared to the Vancouver Island coastal areas during a major wind storm on Dec 15, 2006. Generally, surge amplitudes are found to be greater within the Strait of Georgia when compared to amplitudes in the Strait of Juan de Fuca.
\end{abstract}

@@

The modelled domain extends from the Strait of Juan de Fuca to Puget Sound to Johnstone Strait (Figure \ref{fig:domain}). Bathymetry from the Cascadia physiography data set \citep{haugerud1999digital} was smoothed to limit large changes in depth across grid cells, so that $\Delta h/\bar{h} \leq 0.8$, where $\Delta h$ is the difference in depth between two adjacent grid cells and $\bar{h}$ is their average depth. For model stability, additional smoothing at the Strait of Juan de Fuca western boundary was imposed to achieve constant depth across the first ten grid cells {\color{red} inward from the open boundary. Depths between 0 m and 4 m were set to 4 m and a minimum depth of 4 m was by NEMO.}  As depicted in Figure \ref{fig:domain}, the numerical grid is rotated $29^{\circ}$ counter-clockwise of North in order to maintain computational efficiency since currents within the Strait of Georgia are mainly aligned with this rotated axis. {\color{red} A magnified image of the complex passages through the San Jaun and Gulf Islands is provided in Figure \ref{fig:zoom}.}

@@

Forcing the model with only eight tidal constituents leads to an error in modelled sea surface height due to the omission of the next leading order tidal constituents such as J$_1$. Comparisons between full tidal predictions and tidal predictions using eight constituents suggest that this error can approach {\color{red} 20} cm during times of high tidal range. {\color{red} As such, model water level predictions were corrected by accounting for this error.Note that the model will contain tidal energy from shallow water constituents due nonlinear interactions in the model equations. As such, the correction is calculated in the following way:  For each storm surge location of interest, two tidal predictions are calculated: 1) A tidal prediction with all constituents excluding long period constituents (SA, SSA, etc), shallow water constituents, and constituents with signal to noise ratio less than two, and 2) a tidal prediction with only the eight constituents used to force the model.  The difference between these two predictions is added to the model ses surface height as a correction. Note that the harmonic analysis to generate the constituents is applied in the same manner as that used in the forcing. That is, the time series from the year of the storm surge is filtered to remove periods of large non-tidal energy and then the harmonic analysis is performed. The total modelled water level is determined by adding the mean sea level from the tide gauge record over 1983-2001 to the corrected model sea surface height.} This correction is only applied to the modelled water level and not the modelled residuals. All times are reported in UTC.

@@

The model's skill is reported through a few statistical measures of the the total water level $\eta$ and the residual $\delta$. First, the mean absolute error ($\bar{e}$), defined as
\begin{equation}
\bar{e} = \text{mean}\left(\left| M - O \right|\right),
\end{equation}
where $M$ and $O$ are the modelled and observed quantities of interest, is a common measure that estimates the average error in the model. Next, a measure comparing the variance of the error to the variance of the observations defined as \citep{thompson2003prediction},
\begin{equation}
\gamma^2 = \frac{\text{var}\left(M-O\right)}{\text{var}\left(O\right)},
\end{equation}
is also calculated. Smaller values of $\gamma^2$ indicate better model skill. This quantity was used by \citet{bernier2006predicting} and \citet{bernier2010tide} in storm surge predictions in Eastern Canada and is calculated here as a point of comparison. A third measure, called the Willmott skill score \citep{willmott1982some}, is also used to assess overall model skill in predicting the water level and residual. This quantity is defined as
\begin{equation}
WS = 1 - \frac{\sum_{i=1}^N \left(M_i - O_i\right)^2}{\sum_{i=1}^N \left(|M_i'| + |O_i'|\right)^2},
\end{equation} 
where $M_i' = M_i-\bar{O}$ and $O_i'=O_i-\bar{O}$ and $\bar{O}$ is the mean of the observations. This score is bounded by zero and one and values close to one indicate good model skill. Since the Willmott skill score is nondimensional, it is an appropriate statistic to report when comparing skills of different variables, in this case, the total water level and the residuals. Since the total water level has a larger variability than the residuals, the $\gamma^2$ measure should be much lower for the total water levels; hence, the Willmott skill score provides an additional point of comparison. These three statistics, along with the mean observed values and mean modelled values (which indicate the model's bias) are reported in Table \ref{tab:statistics} for each hindcast.

@@

A more detailed evaluation of the model performance is provided by a group of statistical measures, including the mean absolute error, the observed mean, the modelled mean, $\gamma^2$, and the Willmott skill score for both the water level and residuals (Table \ref{tab:statistics}). The statistics are calculated with hourly output over the three-day period displayed in Figure \ref{fig:feb2006}. The mean absolute error in the water level ranges from 7.6 cm at Victoria to 13.1 cm at Campbell River, whereas, the mean absolute error in the residuals is smaller at all stations; however, this measure does not account for the large variability in water levels due to the tides. A statistic that does take into account this variability is $\gamma^2$ which suggests very good agreement in both the predicted water levels and residuals. Good agreement is also reflected in high Willmott skill scores at all locations. Note that the $\gamma^2$ measures and Willmott skill scores degrade slightly for the residuals, likely because the well-tuned tidal signal dominates the water level predictions.

@@

An additional storm surge hindcast for a Nov 16, 2006 storm surge is provided as another means of assessing model performance and is mentioned briefly here. The model performs reasonably well in this hindcast; Willmott skill scores are greater than 0.94 (Table \ref{tab:statistics}).  Further, the $\gamma^2$ statistic suggests that the standard deviation of the error in the modelled residuals is about 40\% ($\sqrt{0.16}$) of the standard deviation of the observed residuals. The best match in maximum surge occurred at Campbell River with a 6.2 cm underestimate by the model. {\color{red} The timing of the maximum surge was too early at all stations except Campbell River.}

>>>>

\begin{abstract}
The Strait of Georgia is a large, semi-enclosed body of water between Vancouver Island and the mainland of British Columbia connected to the Pacific Ocean via the Strait of Juan de Fuca at the south and Johnstone Strait at the north. During the winter months, coastal communities in the Strait of Georgia are at risk to flooding caused by storm surges, a natural hazard that can occur when a strong storm coincides with high tide. This investigation produces storm surge hindcasts using a three-dimensional numerical ocean model for the Strait of Georgia and the surrounding bodies of water (Strait of Juan de Fuca, Puget Sound, and Johnstone Strait) collectively known as the Salish Sea. The numerical model employs the Nucleus for European Modelling of the Ocean (NEMO) architecture in a regional configuration. The model is evaluated through {\color{red} the} comparisons of tidal elevation harmonics and storm surge with observations. Important forcing factors contributing to storm surges have been assessed. It is shown that the surge entering the domain from the Pacific Ocean contributes most significantly to the {\color{red} surge amplitude within the Strait of Georgia. Comparisons between simulations with high resolution and low resolution atmospheric forcing further emphasize that remote forcing is the dominant factor in surge amplitudes of this region.} In addition, local wind patterns cause a slight increase in surge amplitude on the mainland side of the Strait of Georgia when compared to the Vancouver Island coastal areas during a major wind storm on Dec 15, 2006. Generally, surge amplitudes are found to be greater within the Strait of Georgia when compared to amplitudes in the Strait of Juan de Fuca.
\end{abstract}

@@

The modelled domain extends from the Strait of Juan de Fuca to Puget Sound to Johnstone Strait (Figure \ref{fig:domain}). Bathymetry from the Cascadia physiography data set \citep{haugerud1999digital} was smoothed to limit large changes in depth across grid cells, so that $\Delta h/\bar{h} \leq 0.8$, where $\Delta h$ is the difference in depth between two adjacent grid cells and $\bar{h}$ is their average depth. For model stability, additional smoothing at the Strait of Juan de Fuca western boundary was imposed to achieve constant depth across the first ten grid cells {\color{red} inward from the open boundary. Depths between 0 m and 4 m were set to 4 m and a minimum depth of 4 m was imposed by NEMO.}  As depicted in Figure \ref{fig:domain}, the numerical grid is rotated $29^{\circ}$ counter-clockwise of North in order to maintain computational efficiency since currents within the Strait of Georgia are mainly aligned with this rotated axis. {\color{red} A magnified image of the complex passages through the San Juan and Gulf Islands is provided in Figure \ref{fig:zoom}.}

@@

Forcing the model with only eight tidal constituents leads to an error in modelled sea surface height due to the omission of the next leading order tidal constituents such as J$_1$. Comparisons between full tidal predictions and tidal predictions using eight constituents suggest that this error can approach {\color{red} 20} cm during times of high tidal range. {\color{red} As such, model water level predictions were corrected by accounting for this error. Note that the model will contain tidal energy from shallow water constituents due nonlinear interactions in the model equations. As such, the correction is calculated in the following way:  For each storm surge location of interest, two tidal predictions are calculated: 1) A tidal prediction with all constituents excluding long period constituents (SA, SSA, etc), shallow water constituents, and constituents with signal to noise ratio less than two, and 2) a tidal prediction with only the eight constituents used to force the model.  The difference between these two predictions is added to the model ses surface height as a correction. Note that the harmonic analysis to generate the constituents is applied in the same manner as that used in the forcing. That is, the time series from the year of the storm surge is filtered to remove periods of large non-tidal energy and then the harmonic analysis is performed. The total modelled water level is determined by adding the mean sea level from the tide gauge record over 1983-2001 to the corrected model sea surface height.} This correction is only applied to the modelled water level and not the modelled residuals. All times are reported in UTC.

@@

The model's skill is reported through a few statistical measures of the the total water level $\eta$ and the residual $\delta$. First, the mean absolute error ($\bar{e}$), defined as
\begin{equation}
\bar{e} = \text{mean}\left(\left| M - O \right|\right),
\end{equation}
where $M$ and $O$ are the modelled and observed quantities of interest, is a common measure that estimates the average error in the model. Next, a measure comparing the variance of the error to the variance of the observations defined as \citep{thompson2003prediction},
\begin{equation}
\gamma^2 = \frac{\text{var}\left(M-O\right)}{\text{var}\left(O\right)},
\end{equation}
is also calculated. Smaller values of $\gamma^2$ indicate better model skill. This quantity was used by \citet{bernier2006predicting} and \citet{bernier2010tide} in storm surge predictions in Eastern Canada and is calculated here as a point of comparison. A third measure, called the Willmott skill score \citep{willmott1982some}, is also used to assess overall model skill in predicting the water level and residual. This quantity is defined as
\begin{equation}
WS = 1 - \frac{\sum_{i=1}^N \left(M_i - O_i\right)^2}{\sum_{i=1}^N \left(|M_i'| + |O_i'|\right)^2},
\end{equation} 
where $M_i' = M_i-\bar{O}$ and $O_i'=O_i-\bar{O}$ and $\bar{O}$ is the mean of the observations. This score is bounded by zero and one and values close to one indicate good model skill. Since the Willmott skill score is nondimensional, it is an appropriate statistic to report when comparing skills of different variables, in this case, the total water level and the residuals. Since the total water level has a larger variability than the residuals, the $\gamma^2$ measure should be much lower for the total water levels; hence, the Willmott skill score provides an additional point of comparison. These three statistics, along with the mean observed values and mean modelled values (which indicate the model's bias) are reported in Table \ref{tab:statistics} for each hindcast.  {\color{red} Observed and modelled maximum water levels and surge values as well as differences in their timing are reported in Table \ref{tab:peak}.}

@@

A more detailed evaluation of the model performance is provided by a group of statistical measures, including the mean absolute error, the observed mean, the modelled mean, $\gamma^2$, and the Willmott skill score for both the water level and residuals (Table \ref{tab:statistics}). The statistics are calculated with hourly output over the three-day period displayed in Figure \ref{fig:feb2006}. The mean absolute error in the water level ranges from {\color{red}8.2 cm} at Victoria to {\color{red}12.2} cm at Campbell River, whereas, the mean absolute error in the residuals is smaller at all stations; however, this measure does not account for the large variability in water levels due to the tides. A statistic that does take into account this variability is $\gamma^2$ which suggests very good agreement in both the predicted water levels and residuals. Good agreement is also reflected in high Willmott skill scores at all locations. Note that the $\gamma^2$ measures and Willmott skill scores degrade slightly for the residuals, likely because the well-tuned tidal signal dominates the water level predictions.

@@

An additional storm surge hindcast for a Nov 16, 2006 storm surge is provided as another means of assessing model performance and is mentioned briefly here. The model performs reasonably well in this hindcast; Willmott skill scores are greater than {\color{red}0.95} (Table \ref{tab:statistics}).  Further, the $\gamma^2$ statistic suggests that the standard deviation of the error in the modelled residuals is about 40\% ($\sqrt{0.16}$) of the standard deviation of the observed residuals. The best match in maximum surge occurred at Campbell River with a {\color{red}4.3 cm} underestimate by the model. {\color{red} The timing of the maximum surge was too early at all stations except Campbell River.}
